---
title: "Modeling"
author: "Lukas Nick"
date: "30. Oktoober 2015"
output: html_document
---

```{r}
setwd('Documents/ml/datascience/capstone-project/')

# load libraries
library(ggplot2)
source('utils.R')
source('db.R')
source('multiplot.R')
```


# Exploratory Modeling

## Question 1: Is the star-rating a reviewer gives influenced by the current average star rating a business has got at the time of the new review?

### Is there a correlation between the new rating and the average of ratings so far?

* show correlation plot with those correlations filtered out that have fewer than 90 cases
* answer: yes, there is a correlation

### Control for the "general influence" of the business

* create 2 new variables: 
  1. lastn: average rating of the last 10 (20?) reviews (or within last month? no: could be only very few)
  2. baserate: average of all older ratings (from 0 to now - 10)
* use these two variables 

```{r}

NUM.RECENT.RATINGS = 10

sql <- "SELECT r.business_id, 
        array_to_string( array_agg( r.date || ';' || r.stars || ';' || r.useful ), ',') as ratings, 
        count(r.id) as num_reviews
        FROM (SELECT business_id, date, stars, (votes->>'useful')::text::int as useful, id 
              FROM reviews 
              ORDER BY business_id, date ASC) r 
        GROUP BY r.business_id
        ORDER BY r.business_id"
agg.ratings <- query.result(sql)

# extract the information stored in the ratings field and created separate fields
calculate_avgs <- function(rating_string) {
  ratings <- (strsplit(rating_string, ',', fixed=T))[[1]]
  num.ratings <- length(ratings)
  base.range <- 1 : (num.ratings-NUM.RECENT.RATINGS)
  recent.range <- (num.ratings-NUM.RECENT.RATINGS+1) : num.ratings
  base.ratings <- ratings[base.range]
  recent.ratings <- ratings[recent.range]
  base.avg   <- mean(sapply(base.ratings,   function(rtg){ return(as.numeric( strsplit(rtg, ';')[[1]][2] )) }))
  recent.avg <- mean(sapply(recent.ratings, function(rtg){ return(as.numeric( strsplit(rtg, ';')[[1]][2] )) }))
  last.rating <- as.numeric( strsplit( ratings[ length(ratings) ], ';')[[1]][2] )  # stars of latest rating
  useful      <- as.numeric( strsplit( ratings[ length(ratings) ], ';')[[1]][3] )  # usefulness of latest rating
  return( cbind(base.avg, recent.avg, last.rating, useful) )
}

# drop businesses with fewer then 2xNUM.RECENT.RATINGS
agg.ratings <- agg.ratings[ agg.ratings$num_reviews > 2*NUM.RECENT.RATINGS, ]
agg.ratings[,4:7] <- t(sapply( agg.ratings$ratings, calculate_avgs ))
names(agg.ratings)[4:7] <- cbind("base.avg.rating", "recent.avg.rating", "last.rating", "useful")

  ```

Correlate current rating with base-rating and recent ratings

Plot base avg. rating and recent avg rating per stars (last-rating):
```{r}
p1 <- ggplot(agg.ratings, aes(last.rating, base.avg.rating)) + geom_point(alpha=1/100) 
p2 <- ggplot(agg.ratings, aes(last.rating, recent.avg.rating)) + geom_point(alpha=1/100)
multiplot(p1, p2, cols = 2)
```

```{r}
library(Hmisc)
# use Spearman correlation since last.rating isn't normally distributed at all
rcorr(as.matrix(agg.ratings[,c("base.avg.rating", "recent.avg.rating", "last.rating")]), type="spearman")
```

Single regression of average ratings of the business on the last.rating:
```{r}
lm.stars.base <- lm(agg.ratings$last.rating ~ agg.ratings$base.avg.rating)
summary( lm.stars.base )
```

Single regression of average recent ratings on the last.rating:
```{r}
summary( lm(agg.ratings$last.rating ~ agg.ratings$recent.avg.rating))
```

Multiple regression to control for the "business base rate" (captured by "base.avg.rating"):
```{r}
summary( lm(agg.ratings$last.rating ~ agg.ratings$base.avg.rating + agg.ratings$recent.avg.rating) )
```

Is the base.avg.rating term even necessary?
```{r}
fit1 <- lm( last.rating ~ recent.avg.rating, data=agg.ratings)
fit2 <- update( fit1, lm( last.rating ~ recent.avg.rating + base.avg.rating, data=agg.ratings) )
anova(fit1, fit2)

# and the other way?
fit1r <- lm( last.rating ~ base.avg.rating, data=agg.ratings)
fit2r <- update( fit1r, lm( last.rating ~ base.avg.rating + recent.avg.rating, data=agg.ratings) )
anova(fit1r, fit2r)

```

Using Multiple Regression is not really correct because the outcome variable, the stars rating, is discrete, not normally distributed, and can go only from 1 to 5.
More appropriate would be a ordered logistic regression.

```{r}
# Following http://www.ats.ucla.edu/stat/r/dae/ologit.htm

# First add a new variable to agg.ratings that is just last.rating converted to a factor, since polr requires this
agg.ratings$stars <- as.factor(agg.ratings$last.rating)
library(MASS)
m <- polr(stars ~ base.avg.rating + recent.avg.rating, data = agg.ratings, Hess=TRUE)
summary(m)

# Calculate p-values for the coefficients:
coefficients <- coef(summary(m))

# calculate and store p values
p <- pnorm(abs(coefficients[, "t value"]), lower.tail = FALSE) * 2
# combined table
(coefficients <- cbind(coefficients, "p value" = p))
# Get confidence intervals for the coefficients: 
(ci <- confint(m)) # default method gives profiled CIs
```

So the confidence interval for the coefficient of base.avg.rating includes 0, we cannot consider it different from 0. 
The coefficient of recent.avg.rating though seems to be different from 0.

This says that for a one unit increase in the recent.avg.rating, we can assume a 1.5 increase in the expected value of last.rating on the log odds scale, //if// base.avg.rating is kept constant.

Convert the log odds ratios to the easier interpretable Odds Ratios:
```{r}

# odds ratios
exp(coef(m))
# Confidence intervals for the odds ratios
exp(cbind(OR = coef(m), ci))
```

Since only the coefficient of recent.avg.rating seems to significant, let's only look at this one:

If the value of recent.avg.rating moves one unit (e.g. from 1 to 2), the odds of moving from a 1-star last.rating to a higher rating is multiplied by roughly 4.3. (??)

## Model assumption testing (exclude this)

The ordered logit regression assumes that the "steps" between 1 and 2 stars, 2 and 3 stars, etc, are equal (the "proportional odds assumption"), so there is only one coefficient. Graph Is this assumption appropriate? 

```{r}
sf <- function(y) {
  c('Y>=1' = qlogis(mean(y >= 1)),
    'Y>=2' = qlogis(mean(y >= 2)),
    'Y>=3' = qlogis(mean(y >= 3)),
    'Y>=4' = qlogis(mean(y >= 4)),
    'Y>=5' = qlogis(mean(y >= 5)))
}

(s <- with(agg.ratings, summary(as.numeric(stars) ~ base.avg.rating + recent.avg.rating, fun=sf)))

# 

glm(I(as.numeric(stars) >= 2) ~ base.avg.rating, family="binomial", data = agg.ratings)

```




## Other regression factors, from businesses

Possible factors:

* business category: why should there be a difference?
* city: too many cities, many of them with too few businesses + same question as w/ categories
* state: similar problem; therefore:
* country; but see categories
* the number of reviews a business already has  <-- ONLY THIS ONE

## Are ratings that differ from the recent.avg.rating voted as more 'useful'?
  
* calculate difference of new rating and recent.avg.rating in terms of stdev
* correlate with 'useful'ness


```{r}
```
