---
title: "Explore"
author: "Lukas Nick"
date: "16. Oktoober 2015"
output: html_document
---

```{r}
# Preparatory work

# Some links:
# - All categories: https://www.yelp.com/developers/documentation/v2/all_category_list

setwd('Documents/ml/datascience/capstone-project/')

# load libraries
library(ggplot2)
source('utils.R')
source('db.R')

# Load businesses table
business_sql = "select id, business_id, name, full_address, city, state, 
                CASE WHEN state = ANY('{\"BW\",\"RP\"}'::text[]) THEN 'DE'
                     WHEN state = ANY('{\"EDH\",\"ELN\",\"FIF\",\"HAM\",\"KHL\",\"MLN\",\"SCB\",\"XGL\"}'::text[]) THEN 'GB'
                     WHEN state = ANY('{\"ON\",\"QC\"}'::text[]) THEN 'CA'
                     ELSE 'US' END as country,
                longitude, latitude, stars, review_count, 
                attribs->'By Appointment Only'::text as appointment,
                attribs->'Alcohol'::text as alcohol,
                attribs->'Noise Level'::text as noise,
                attribs->'Has TV'::text as tv,
                attribs->'Attire'::text as attire,
                attribs->'Good for Kids'::text as for_kids,
                attribs->'Good for Groups'::text as for_groups,
                attribs->'Outdoor Seating'::text as outdoor,
                attribs->'Price Range'::text as price_range,
                attribs->'Take-out'::text takeout,
                attribs->'Drive-Thru'::text drivethrou,
                attribs->'Happy Hour'::text as happyhour,
                attribs->'Takes Reservations'::text as reservations,
                attribs->'Waiter Service'::text as waiter_service,
                attribs->'Caters'::text as caters,
                attribs->'Smoking'::text as smoking,
                attribs->'Wheelchair Accessible'::text as wheelchair,
                attribs->'Coat Check'::text as coat_check,
                attribs->'Dogs Allowed'::text as dogs,
                attribs->'Hot and New'::text as hotnew,
                attribs->'Delivery'::text as delivery,
                attribs->'Wi-Fi'::text as wifi,
                attribs->'Good For'->>'dessert'::text as good_for_dessert,
                attribs->'Good For'->>'latenight'::text as good_for_latenight,
                attribs->'Good For'->>'lunch'::text as good_for_lunch,
                attribs->'Good For'->>'dinner'::text as good_for_dinner,
                attribs->'Good For'->>'brunch'::text as good_for_brunch,
                attribs->'Good For'->>'breakfast'::text as good_for_breakfest,
                attribs->'Ambience'->>'romantic'::text as ambiance_romantic,
                attribs->'Ambience'->>'intimate'::text as ambiance_intimate,
                attribs->'Ambience'->>'classy'::text as ambiance_classy,
                attribs->'Ambience'->>'classy'::text as ambiance_classy,
                attribs->'Ambience'->>'hipster'::text as ambiance_hipser,
                attribs->'Ambience'->>'divey'::text as ambiance_divey,
                attribs->'Ambience'->>'touristy'::text as ambiance_touristy,
                attribs->'Ambience'->>'trendy'::text as ambiance_trendy,
                attribs->'Ambience'->>'upscale'::text as ambiance_upscale,
                attribs->'Ambience'->>'casual'::text as ambiance_casual,
                attribs->'Parking'->>'garage'::text as parking_garage,
                attribs->'Parking'->>'street'::text as parking_street,
                attribs->'Parking'->>'validated'::text as parking_validated,
                attribs->'Parking'->>'lot'::text as parking_lot,
                attribs->'Parking'->>'valet'::text as parking_valet,
                attribs->'Music'->>'dj'::text as music_dj,
                attribs->'Music'->>'video'::text as music_video,
                attribs->'Music'->>'jukebox'::text as music_jukebox,
                attribs->'Music'->>'live'::text as music_live,
                attribs->'Music'->>'background_music'::text as music_background,
                attribs->'Music'->>'karaoke'::text as music_karaoke,
                open,
                ARRAY(select name from categories c where c.business_id = b.business_id) as categories
                from businesses b"
buss.rs = query(business_sql)
buss = dbFetch(buss.rs, -1)
```

# Exploratory Analysis

Tasks:

1. Explore the relationships between different features in each data file.
2. Try linking data files together and explore the relationships between features across data files.
3. Identify interesting outcomes that you may want to predict as part of a prediction question / problem
4. Characterize any missing data that may be present in each of the files
5. Many features incorporate free-text data that may need to be parsed, summarized, or quantified in some way. What is the best way to handle these data?

## 1. Relationships between features in each data file

### 1.1. Relationsships in the Business file

#### Potential features:

* City
* State
* Longitude/Latitude
* Stars
* Review_count
* Neighborhoods
* Attributes
* Categories (restaurant, bar, )
* Open

#### Review_count-Stars relationship

```{r}
p1 <- qplot(stars, data=buss, geom="bar")
p2 <- qplot(stars, review_count, data=buss, color=country)
multiplot(p1, p2, cols=2)

qplot(stars, review_count, data=buss, geom="jitter")
```


## Influences on the reviews' star ratings

Questions

1. Is the star-rating a reviewer gives influenced by the current average star rating a business has got at the time of the new review?
2. Is this influence stronger than the influence of other factors, as measured by multiple regression; factors:
  - the city
  - neighboorhoods
  - various attributes; candidates?
  - the number of reviews a business already has
3. Are ratings that _differ_ from the avg (as for example measured by standard deviations) rated as more "useful", or maybe "cool"?  

3. **Alternative question:** Influence of outstanding (negative or positive) ratings on following ratings
  - depending on the outstanding rating's attributes (votes) and its user's compliments 

## Explore questions 1 and 2

Get a list (dataframe) of all reviews ordered by business and review date. For each review, also show the average rating of the business and the number of reviews for this business. 
```{r}

sql <- "SELECT r.business_id, r.review_id, r.user_id, r.date, r.stars,
        (r.votes->'funny')::text::int as funny, 
        (r.votes->'useful')::text::int as useful, 
        (r.votes->'cool')::text::int as cool,
          (SELECT ARRAY[ avg(r2.stars), count(r2.id) ] 
           FROM reviews r2
           WHERE r2.business_id = r.business_id
           AND r2.date < r.date) AS avg_count_sofar
        FROM reviews r
        ORDER BY r.business_id, r.date"    
ratings <- query.result(sql)

# Split avg_count_sofar strings into to separate columns
avg_count <- strsplit( gsub("[{}]", '', ratings[,'avg_count_sofar']), ',' )
avgs   <- sapply(avg_count, function(cv){ return( as.numeric(cv[1]) ) })
counts <- sapply(avg_count, function(cv){ return( as.numeric(cv[2]) ) })
ratings$avg_rating_sofar  <- avgs
ratings$num_ratings_sofar <- counts

head(ratings)
```

Checkout the correlation between the current rating average and the new rating for the second rating, the third, and so on.

First: There are no missings in for the stars and their averages/counts (except that the first review doesn't have an average of the previous ratings):
```{r}
num.missing.stars <- nrow( ratings[is.na(ratings$stars), ])
num.missing.avg.stars.sofar <- nrow( ratings[is.na(ratings$avg_rating_sofar) && ratings$num_ratings_sofar>0, ])
num.missing.num.ratings.sofar <- nrow( ratings[is.na(ratings$num_ratings_sofar), ])

sprintf("num.missing.stars: %d", num.missing.stars)
sprintf("num.missing.avg.stars.sofar: %d", num.missing.avg.stars.sofar)
sprintf("num.missing.num.ratings.sofar: %d", num.missing.num.ratings.sofar)
```

Correlations of new rating with average of previous ones, depending on the number of the ratings (second, third, forth, etc):
```{r}
rating_nums <- 2:max(ratings$num_ratings_sofar)
cors <- sapply(rating_nums, function(i){
  current.sofar <- ratings[ratings$num_ratings_sofar == i, c('stars', 'avg_rating_sofar')]
  return( cor(current.sofar$stars, current.sofar$avg_rating_sofar))
})
plot(rating_nums, cors)

rating.cors <- cbind(rating_nums, cors)
write.table(rating.cors, "rating_correlations.csv", sep=";", row.names = FALSE)  # save for later
```

There seems to be correlation, roughly around 0.4. The variance in the correlations grows with the number of reviews, the tendency seems to be the more reviews, the higher the correlations. Let's put a regression line into this plot:
```{r}

p <- ggplot(rating.cors, aes(rating_nums, cors, colour = cut))
