---
title: "Exploratory Modeling"
author: "Lukas Nick"
date: "30. Oktoober 2015"
output: html_document
---

```{r}
setwd('Documents/ml/datascience/capstone-project/')

# load libraries
library(ggplot2)
source('utils.R')
source('db.R')
source('multiplot.R')
```

# Questions

1. Is the star-rating a reviewer gives influenced by the current average star rating a business has got at the time of the new review?
2. If there is an influence of earlier ratings on a new rating, is it bigger than the influence of the number of reviews at the time of the new rating?
3. Are ratings that differ from the average rating for this business rated more or less 'useful'?

# Modeling

## Question 1: Is the star-rating a reviewer gives influenced by the current average star rating a business has got at the time of the new review?

### Is there a correlation between the new rating and the average of ratings so far?

* show correlation plot with those correlations filtered out that have fewer than 90 cases
* answer: yes, there is a correlation

### Control for the "general influence" of the business

* create 2 new variables: 
  1. lastn: average rating of the last 10 (20?) reviews (or within last month? no: could be only very few)
  2. baserate: average of all older ratings (from 0 to now - 10)
* use these two variables 

```{r}

NUM.RECENT.RATINGS = 20

sql <- "SELECT rb.business_id, 
        array_to_string( array_agg( rb.date || ';' || rb.stars || ';' || rb.useful || ';' || rb.country ), ',') as ratings, 
        count(rb.id) as num_reviews    
        FROM (SELECT r.business_id, r.date, r.stars, (votes->>'useful')::text::int as useful, b.state, r.id,
              CASE WHEN state IN ('BW', 'RP') THEN 'DE'             
                   WHEN state IN ('EDH', 'ELN', 'FIF','HAM', 'KHL', 'MLN', 'SCB','XGL') THEN 'GB'
                   WHEN state IN ('ON','QC') THEN 'CA'
                   ELSE 'US' END as country
              FROM reviews r INNER JOIN businesses b USING (business_id) 
              ORDER BY business_id, date ASC) rb 
        GROUP BY rb.business_id
        ORDER BY rb.business_id"
agg.ratings <- query.result(sql)

# extract the information stored in the ratings field and created separate fields
calculate_avgs <- function(rating_string) {
  ratings <- (strsplit(rating_string, ',', fixed=T))[[1]]
  num.ratings <- length(ratings)
  base.range <- 1 : (num.ratings-NUM.RECENT.RATINGS-1)
  recent.range <- (num.ratings-NUM.RECENT.RATINGS) : (num.ratings -1)
  base.ratings <- ratings[base.range]
  recent.ratings <- ratings[recent.range]
  base.avg   <- mean(sapply(base.ratings,   function(rtg){ return(as.numeric( strsplit(rtg, ';')[[1]][2] )) }))
  recent.avg <- mean(sapply(recent.ratings, function(rtg){ return(as.numeric( strsplit(rtg, ';')[[1]][2] )) }))
  recent.sd  <- sd(  sapply(recent.ratings, function(rtg){ return(as.numeric( strsplit(rtg, ';')[[1]][2] )) }))
  last.rating <- as.numeric( strsplit( ratings[ num.ratings ], ';')[[1]][2] )  # stars of latest rating
  useful      <- as.numeric( strsplit( ratings[ num.ratings ], ';')[[1]][3] )  # usefulness of latest rating
  country     <- ( strsplit( ratings[1], ';')[[1]][4] )            # country of first rating (the same for all ratings) 
  return( list(base.avg = base.avg, recent.avg = recent.avg, recent.sd = recent.sd, last.rating = last.rating, useful=useful, country=country))
}

# drop businesses with fewer then 2xNUM.RECENT.RATINGS
agg.ratings <- agg.ratings[ agg.ratings$num_reviews > 2*NUM.RECENT.RATINGS, ]
calculated.cols <- t(sapply( agg.ratings$ratings, calculate_avgs ))
row.names(calculated.cols) <- c()  # remove the row names
calculated.df <- as.data.frame(calculated.cols)
agg.ratings[,4:9] <- calculated.df
names(agg.ratings)[4:9] <- cbind("base.avg.rating", "recent.avg.rating", "recent.sd", "last.rating", "useful", "country")
# convert lists to numeric/character:
agg.ratings$base.avg.rating <- as.numeric(agg.ratings$base.avg.rating)
agg.ratings$recent.avg.rating <- as.numeric(agg.ratings$recent.avg.rating)
agg.ratings$recent.sd <- as.numeric(agg.ratings$recent.sd)
agg.ratings$last.rating <- as.numeric(agg.ratings$last.rating)
agg.ratings$useful <- as.numeric(agg.ratings$useful)
agg.ratings$country <- as.character(agg.ratings$country)

```

Correlate current rating with base-rating and recent ratings

Plot base avg. rating and recent avg rating per stars (last-rating):
```{r}
p1 <- ggplot(agg.ratings, aes(last.rating, base.avg.rating)) + geom_point(alpha=1/100) 
p2 <- ggplot(agg.ratings, aes(last.rating, recent.avg.rating)) + geom_point(alpha=1/100)
multiplot(p1, p2, cols = 2)
```

```{r}
library(Hmisc)
# use Spearman correlation since last.rating isn't normally distributed at all
rcorr(as.matrix(agg.ratings[,c("base.avg.rating", "recent.avg.rating", "last.rating")]), type="spearman")
```

Multiple regression to control for the "business base rate" (captured by "base.avg.rating"):
```{r}
summary( lm(agg.ratings$last.rating ~ agg.ratings$base.avg.rating + agg.ratings$recent.avg.rating) )
```

Is the recent.avg.rating term even necessary?
```{r}
fit1r <- lm( last.rating ~ base.avg.rating, data=agg.ratings)
fit2r <- update( fit1r, lm( last.rating ~ base.avg.rating + recent.avg.rating, data=agg.ratings) )
anova(fit1r, fit2r)

```

-> Yes.

Using Multiple Regression is not really correct because the outcome variable, the stars rating, is discrete, not normally distributed, and can go only from 1 to 5.
More appropriate would be a ordered logistic regression.

```{r}
# Following http://www.ats.ucla.edu/stat/r/dae/ologit.htm

# First add a new variable to agg.ratings that is just last.rating converted to a factor, since polr requires this
agg.ratings$stars <- as.factor(agg.ratings$last.rating)
library(MASS)
m <- polr(stars ~ base.avg.rating + recent.avg.rating, data = agg.ratings, Hess=TRUE)
summary(m)

# Calculate p-values for the coefficients:
coefficients <- coef(summary(m))

# calculate and store p values
p <- pnorm(abs(coefficients[, "t value"]), lower.tail = FALSE) * 2
# combined table
(coefficients <- cbind(coefficients, "p value" = p))
# Get confidence intervals for the coefficients: 
(ci <- confint(m)) # default method gives profiled CIs
```

So both the 'base rating' and the recent avg rating 'influence a new rating: the confidence interval for both variables doesn't include 0.

This says that for a one unit increase in the recent.avg.rating, we can assume a 1.5 increase in the expected value of last.rating on the log odds scale, //if// base.avg.rating is kept constant.

Convert the log odds ratios to the easier interpretable Odds Ratios:
```{r}

# odds ratios
exp(coef(m))
# Confidence intervals for the odds ratios
exp(cbind(OR = coef(m), ci))
```

If the value of base.avg.rating moves one unit (e.g. from 1 to 2), the odds of moving from a 1-star last.rating to a higher rating is multiplied by 1.8
Similarily, if the value of recent.avg.rating moves one unit (e.g. from 1 to 2), the odds of moving from a 1-star last.rating to a higher rating is roughly doubled.


## Influence of country and the number of reviews at the time of the last rating


```{r}
table(agg.ratings$country)

m2 <- polr(stars ~ base.avg.rating + recent.avg.rating + num_reviews + country, data = agg.ratings[ agg.ratings$country != 'DE', ], Hess=TRUE)
summary(m2)

# Calculate p-values for the coefficients:
coefficients2 <- coef(summary(m2))

# calculate and store p values
p2 <- pnorm(abs(coefficients2[, "t value"]), lower.tail = FALSE) * 2
# combined table
(coefficients2 <- cbind(coefficients2, "p value" = p2))
# Get confidence intervals for the coefficients: 
(ci2 <- confint(m2)) # default method gives profiled CIs

# odds ratios
exp(coef(m2))
# Confidence intervals for the odds ratios
exp(cbind(OR = coef(m2), ci2))

```

So number of reviews and country don't affect the rating, and the relations between both the base average rating and recent average rating with a new rating hold, even if we control for the number of reviews and country.

## Are ratings that differ from the recent.avg.rating voted as more 'useful'?
  
* calculate difference of new rating and recent.avg.rating in terms of stdev
* correlate with 'useful'ness


```{r}
valid.sds <- which(agg.ratings$recent.sd > 0)
agg.ratings.with.sd <- agg.ratings[ valid.sds, ]
agg.ratings.with.sd$deviation <- abs(agg.ratings.with.sd$recent.avg.rating - agg.ratings.with.sd$last.rating) / agg.ratings.with.sd$recent.sd

cor(agg.ratings.with.sd$deviation, agg.ratings.with.sd$useful)
```

There is no correlation between the deviation of the last rating form the recent average and the 'usefulness', i.e. the number of times a review got the vote 'useful'.
