---
title: "Influence of Recent Ratings on New Ratings in Yelp Business Reviews"
author: "Lukas Nick"
references:
- author:
  - family: Asch
    given: S. E.
  container-title: G. E. Swanson, T. M. Newcomb & E. L. Hartley (Eds.), Readings in
    social psychology (2nd ed., pp. 2-11).
  id: aesch
  issued:
    year: 1952
  publisher: New York:NY Holt
  title: Effects of group pressure on the modification and distortion of judgements
  type: book
- URL: https://en.wikipedia.org/wiki/Ordered_logit
  author:
  - family: Wikipedia
  id: wikipedia
  title: Ordered Logit
- URL: http://www.ats.ucla.edu/stat/r/dae/ologit.htm
  author:
  - family: idre
    given: UCLA
  id: ologit
  title: R Data Analysis Examples - Ordinal Logistic Regression
output: pdf_document
---

```{r, echo=FALSE, include=FALSE}

# Setup
setwd('~/Documents/ml/datascience/capstone-project/')

# load libraries
library(ggplot2)
source('utils.R')
source('db.R')
source('multiplot.R')
library(Hmisc)
library(MASS)
```


# Title

This study evaluates whether a new star rating is related to the average of older ratings. It also considers whether this relationship depends on the country or the number of reviews at the time of the new rating. 

Furthermore, it addresses the question whether ratings that deviate from the average of the earlier ratings are more often voted as 'useful' than non deviating ratings.

The first two questions are tackled using ordered logit models, the third question simply calculates a correlation.

# Introduction

Most users of Yelp, when they review a business and judge it with a rating (i.e. a number of "stars" between 1 and 5), have already seen the older reviews of that business, since they are displayed on the same page. Social Psychology suggests that other people's opinions influence ones own judgment; see, among many, for example [@aesch]. Therefore it seems reasonable to assume that older ratings, corresponding to other people's options, could influence a user's new rating.

If that was the case, designers of Yelp or similar services that use any kind of rating might want to look into ways of "blind" voting, so that new raters can see other users' ratings only _after_ having stated their own rating.

On obvious problem when comparing previous ratings with new ratings is that we have to assume a "business factor" influencing both previous ratings and older ratings: if the business is, let's say, in general "good", then both previous users and the new user will tend to rate it higher, and vice versa. We therefore need a way to seperate a general "business factor" from a "social factor".

To achieve this, two measures of previous ratings were calculated: the average of the last 20 ratings (named the "recent average rating"), and the average of the rest of the older ratings (named "base average rating"). The base average rating is thought to represent the "general business factor", and the recent average rating is considered the one that has some social influence (assuming that rarely users go further back than one page to look at older ratings, especially on a mobile device).

Those two factors were distinguished with a multivariate model.

# Methods and Data

## Dataset

I used the data from the [Yelp Dataset Challenge](http://www.yelp.com/dataset_challenge): it contains 1.5M reviews of 481 thousend businesses, by 366 thousend users. 

To better handle this amount of data, I programmed a small [Rails application](https://github.com/lunas/yelp_seeder) to seed the data (which is available in JSON format) into a PostgreSQL database.

I then created a table that lists each business in a row, and in separate columns contains:

* the latest rating; the rating of the new user that will be compared to previous ratings
* the recent average rating: the average of the ratings of the most recent 21 reviews (excluding the latest rating)
* the base average rating: the average of all older reviews (i.e. all ratings excluding the latest 21 ratings)
* the number of reviews at the time of the latest rating
* the number of times the latest rating got voted as 'useful'
* the country of the business: either USA, Canada, Great Britain or Germany.

```{r, echo=FALSE, cache=TRUE}

NUM.RECENT.RATINGS = 20

sql <- "SELECT rb.business_id, 
        array_to_string( array_agg( rb.date || ';' || rb.stars || ';' || rb.useful || ';' || rb.country ), ',') as ratings, 
        count(rb.id) as num_reviews    
        FROM (SELECT r.business_id, r.date, r.stars, (votes->>'useful')::text::int as useful, b.state, r.id,
              CASE WHEN state IN ('BW', 'RP') THEN 'DE'             
                   WHEN state IN ('EDH', 'ELN', 'FIF','HAM', 'KHL', 'MLN', 'SCB','XGL') THEN 'GB'
                   WHEN state IN ('ON','QC') THEN 'CA'
                   ELSE 'US' END as country
              FROM reviews r INNER JOIN businesses b USING (business_id) 
              ORDER BY business_id, date ASC) rb 
        GROUP BY rb.business_id
        ORDER BY rb.business_id"
agg.ratings <- query.result(sql)

# extract the information stored in the ratings field and created separate fields
calculate_avgs <- function(rating_string) {
  ratings <- (strsplit(rating_string, ',', fixed=T))[[1]]
  num.ratings <- length(ratings)
  base.range <- 1 : (num.ratings-NUM.RECENT.RATINGS-1)
  recent.range <- (num.ratings-NUM.RECENT.RATINGS) : (num.ratings -1)
  base.ratings <- ratings[base.range]
  recent.ratings <- ratings[recent.range]
  base.avg   <- mean(sapply(base.ratings,   function(rtg){ return(as.numeric( strsplit(rtg, ';')[[1]][2] )) }))
  recent.avg <- mean(sapply(recent.ratings, function(rtg){ return(as.numeric( strsplit(rtg, ';')[[1]][2] )) }))
  recent.sd  <- sd(  sapply(recent.ratings, function(rtg){ return(as.numeric( strsplit(rtg, ';')[[1]][2] )) }))
  last.rating <- as.numeric( strsplit( ratings[ num.ratings ], ';')[[1]][2] )  # stars of latest rating
  useful      <- as.numeric( strsplit( ratings[ num.ratings ], ';')[[1]][3] )  # usefulness of latest rating
  country     <- ( strsplit( ratings[1], ';')[[1]][4] )            # country of first rating (the same for all ratings) 
  return( list(base.avg = base.avg, recent.avg = recent.avg, recent.sd = recent.sd, last.rating = last.rating, useful=useful, country=country))
}

# drop businesses with fewer then 2xNUM.RECENT.RATINGS
agg.ratings <- agg.ratings[ agg.ratings$num_reviews > 2*NUM.RECENT.RATINGS, ]
calculated.cols <- t(sapply( agg.ratings$ratings, calculate_avgs ))
row.names(calculated.cols) <- c()  # remove the row names
calculated.df <- as.data.frame(calculated.cols)
agg.ratings[,4:9] <- calculated.df
names(agg.ratings)[3:9] <- cbind("num.reviews", "base.avg.rating", "recent.avg.rating", "recent.sd", "last.rating", "useful", "country")
# convert lists to numeric/character:
agg.ratings$base.avg.rating <- as.numeric(agg.ratings$base.avg.rating)
agg.ratings$recent.avg.rating <- as.numeric(agg.ratings$recent.avg.rating)
agg.ratings$recent.sd <- as.numeric(agg.ratings$recent.sd)
agg.ratings$last.rating <- as.numeric(agg.ratings$last.rating)
agg.ratings$useful <- as.numeric(agg.ratings$useful)
agg.ratings$country <- as.character(agg.ratings$country)

```

To calculate averages of the latest 20 ratings and meaningful averages of the even older base ratings, we need businesses with at least 40 reviews (2 * 20). Therefore businesses with fewer reviews were filtered out.
```{r, echo=FALSE}
# head(agg.ratings[,c('business_id', 'num.reviews', 'base.avg.rating', 'recent.avg.rating', 'last.rating', 'useful','country', 'recent.sd')])
print(agg.ratings[1:5,c('business_id', 'num.reviews', 'base.avg.rating', 'recent.avg.rating', 'last.rating', 'useful','country', 'recent.sd')], row.names=FALSE)
```

The last column, 'recent.sd', contains the standard deviation of the 20 most recent ratings. It is used to calculate the deviation of the latest rating from the 20 most recent ratings in terms of their standard calculation.
```{r, echo=FALSE}
valid.sds <- which(agg.ratings$recent.sd > 0)
agg.ratings.with.sd <- agg.ratings[ valid.sds, ]
agg.ratings.with.sd$deviation <- abs(agg.ratings.with.sd$recent.avg.rating - agg.ratings.with.sd$last.rating) / agg.ratings.with.sd$recent.sd
agg.ratings.with.sd$diff <- abs(agg.ratings.with.sd$recent.avg.rating - agg.ratings.with.sd$last.rating) 
```

## Data exploration

### Base, recent and new ratings

```{r, echo=FALSE, message=FALSE}
h2 <- qplot(base.avg.rating,   data = agg.ratings, geom = "histogram")
h3 <- qplot(recent.avg.rating, data = agg.ratings, geom = "histogram")
h4 <- qplot(last.rating, data = agg.ratings, geom = "histogram")
multiplot(h2, h4, h3, cols = 2)
```

The distribution of both base.avg.rating and recent.avg.rating is close to a normal. Not so the distribution of the latest ratings.

How are the ratings distributed in the three countries?

Amount of reviews per country:
```{r, echo=FALSE}
table(agg.ratings$country, agg.ratings$last.rating)
```

```{r, echo=FALSE,include=FALSE}
h1 <- qplot(log(last.rating),  data = agg.ratings, geom = "histogram", facets = country ~ .) + scale_y_continuous(trans = "log2")
# h1 
```

The distribution of last.rating seems to be roughly similar in the USA, Canada and Great Britain. There are no german businesses any more: all of them had less than 40 reviews and were therefore filtered out.

Plot the base.avg.rating and recent.avg.rating per stars (last-rating):

```{r, echo=FALSE}
p1 <- ggplot(agg.ratings, aes(last.rating, base.avg.rating)) + geom_point(alpha=1/100) 
p2 <- ggplot(agg.ratings, aes(last.rating, recent.avg.rating)) + geom_point(alpha=1/100)
multiplot(p1, p2, cols = 2)
```

```{r, echo=FALSE}
p <-ggplot(agg.ratings, aes(x = last.rating, y = recent.avg.rating)) +
      geom_boxplot(size = .75) +
      geom_jitter(alpha = .5) +
      facet_grid(country ~ ., margins = TRUE) +
      theme(axis.text.x = element_text(hjust = 1, vjust = 1))
# no need to print this one
```

There seems to be a small correlation between the latest rating and the both the base average rating and the recent average rating.

Let's calculate the Spearman correlation (since the latest rating is not normally distributed):

```{r, echo=FALSE}
rcorr(as.matrix(agg.ratings[,c("base.avg.rating", "recent.avg.rating", "last.rating")]), type="spearman")
```

To have a preliminary look at the interaction of those average ratings with the latest rating, let's do a quick multiple regression that predicts latest.rating by recent average rating and controls for base average rating:

```{r, echo=FALSE}
summary( lm(agg.ratings$last.rating ~ agg.ratings$base.avg.rating + agg.ratings$recent.avg.rating) )
```

Apparently the variable recent.avg.rating does add significantly to the model: it is not redunant even though we included the base average rating.

### Deviation and usefulness

As an additional question, we want to know whether deviating ratings are considered more useful.

Both the 'deviation' (difference between latest rating and recent average rating in terms of standard deviation within the recent averate rating) and the 'difference' (absolute difference between the latest rating and the recent average rating) were calculated. Let's look at their relationship with the number of time a rating was considered 'useful':

```{r, echo=FALSE, message=FALSE}
p1 <- ggplot(agg.ratings.with.sd, aes(deviation, useful)) + geom_point()
p2 <- ggplot(agg.ratings.with.sd, aes(diff, useful)) + geom_point()
multiplot(p1, p2, cols = 2)
```

This looks rather as if less deviation ratings were voted more 'useful'.

## Methods

Since the dependent variable, the latest rating, is discrete, not normally distributed, and only ranges from 1 to 5, multiple regression is not the ideal method. Therefore an ordered logistic regression model was calculated.

### Ordered logistic regression

Ordered logistic regression is a regression model for ordinal dependent variables (as our ``latest.rating`` variable, the "stars" given in a review). It uses maximum likelihood to estimate the coefficients [see @wikipedia; @ologit].

```{r, echo=FALSE}
# Following http://www.ats.ucla.edu/stat/r/dae/ologit.htm

# First add a new variable to agg.ratings that is just last.rating converted to a factor, since polr requires this
agg.ratings$stars <- as.factor(agg.ratings$last.rating)
```

The function ``polr`` from the library MASS was used to estimate the coefficients of the ologit model. To compare the influence of the base average rating and the recent average rating with the influence of the number of reviews at the time of the latest rating, ``num.reviews`` was included in the ordered logit model as well.

```{r}
m <- polr(stars ~ base.avg.rating + recent.avg.rating + num.reviews, data = agg.ratings, Hess=TRUE)
```

The variable ``country`` was not included in the model since exploratory analysis showed that the distribution of CA and USA looks very similar, and both CA and GB have to small cell sizes for the ologit model to be stable or meaningful.

### Deviations of latest rating and 'usefulness'

Finally, to see whether ratings that differ from earlier ratings are voted more often as 'useful', the difference of the latest rating in terms of the standard deviation of the 20 most recent ratings was calculated. This new variable, dubbed simply 'deviation', was then correlation with the number the latest rating got voted as 'useful'. The Spearman correlation was used because of the non-normal distribution of the variable 'useful'.

# Results

## Base average rating, recent average rating, number of reviews and the last rating

```{r, echo=FALSE}
# Following http://www.ats.ucla.edu/stat/r/dae/ologit.htm
#summary(m)

# Calculate p-values for the coefficients:
coefficients <- coef(summary(m))

# calculate and store p values
p <- pnorm(abs(coefficients[, "t value"]), lower.tail = FALSE) * 2
# combined table
(coefficients <- cbind(coefficients, "p value" = p))
```

We see that both the base average rating and the recent average rating exhibit a significant influence on a new rating, whereas the number of reviews seems to be irrelevant.

To easier interpret the coeffient, let's convert them to odds rations and calculate there confidence intervals:

```{r, echo=FALSE, message=FALSE}
# Get confidence intervals for the coefficients: 
ci <- confint(m) # default method gives profiled CIs

# odds ratios
# exp(coef(m))
# Confidence intervals for the odds ratios
exp(cbind(OddsRatios = coef(m), ci))
```

So we can expect an increase in the base average rating of one start to increase the _odds_ for the lastest rating to get a higher star-rating to rise by 82%.

Likewise, an increase in the recent average rating of one star raises the _odds_ for the latest rating to get a (any) higher rating more than twice (2.14).

Both these odds have a confidence intervall way above one, as opposed to the number of reviews: with 95% probability, one more review does not change the odds of the latest review to have any higher star rating.

##  Deviations of latest rating and 'usefulness'

The graphical evaluation in the exploratory analysis hinted at a negative relationship between deviation of latest rating from recent average rating and 'usefulness' of the latest rating. The Spearman method reveals no correlation that could be considered 'significant' in any meaningful way:

```{r, echo=FALSE}
#cor(agg.ratings.with.sd$deviation, agg.ratings.with.sd$useful)
#cor(agg.ratings.with.sd$diff, agg.ratings.with.sd$useful)
rcorr(as.matrix(agg.ratings.with.sd[,c("deviation", "diff", "useful")]), type="spearman")
```

  
# Discussion

The main question of this study is whether the recent ratings have an influence on a new rating and whether this influence is not entirely predicted by the 'base rating' for a business. In other words, even if a business is rated in general as, say, "good", do the recent ratings influence a new rating. To control this 'base rating' of a business, the 20 most recent ratings were separated from all the older ratings, i.e. an average was calculated for both older base rating and the 20 most recent ratings (except the very latest rating). Then a multiple regression model was fitted, using ordered logit regression, to estimate the influence of the recent rating on to the latest rating, controlling for the base rating.

The model coefficients from the fitted model hint at an influence of both the base and the recent average rating on the new (latest) rating. But the recent average rating does have a significant (even bigger) influence on the new rating, even when controlling for the business' base rating. Therefore we can say, yes, recent ratings do influence new ratings, independently of whether how the business is rated in general.

The number of reviews a business has at the time of the latest rating does not seem to have any relevance on a new rating.

There are though a couple of drawbacks of this Ordered Logit Regression:

* this method assumes that the coefficients that describe the relationship between a 1-star rating and any higher rating is the same as between a 1-or-2-star rating and any higher rating (the "proportional odds assumption"). This assumption has not been tested here.
* it is difficult to come up with diagnostics for the model fitted here. I don't know _how_ fit it is.

The strong significance of the coefficients related to the main question of this study indicate though, that the hypothesized relationship of recent ratings and new ratings helds, controlled for the business' base rating.

Of course one should keep in mind, that this data is purely 'correlational', not causal.

A secondary question was studied, namely whether new ratings that diverge from the average of the recent ratings are considered more 'useful' by other Yelp users. Correlational results don't indiciate that there is any such relationship. Visual results hint at the possibility of a non-linear relationship between deviation and 'usefulness' (see section "Deviation and usefulness" in "Data and Methods") which a correlation would fail to discover.


## References